---
title: "assignment 8"
author: "anubrata roy"
date: "2026-03-31"
output: html_document
---

# Part (a): Train-Test Split

Observations from **1990–2008** form the training set, and **2009–2010** form the test set.

```{r}
library(ISLR)
library(dplyr)
library(knitr)
library(kableExtra)

data(Weekly)

# Train: 1990 to 2008
train <- Weekly %>% filter(Year <= 2008)

# Test: 2009 to 2010
test  <- Weekly %>% filter(Year >= 2009)

cat("Training set size:", nrow(train), "\n")
cat("Test set size    :", nrow(test),  "\n")
cat("\nDistribution of Direction in Training Set:\n")
print(table(train$Direction))
cat("\nDistribution of Direction in Test Set:\n")
print(table(test$Direction))
```


# Part (b): Logistic Regression – Training Data

We fit a logistic regression with `Direction` as the binary response and `Lag1`, `Lag2`, `Lag3`, `Lag4`, `Lag5`, and `Volume` as predictors.

```{r}
# Fit logistic regression model
logit_model <- glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
                   data   = train,
                   family = binomial(link = "logit"))

summary(logit_model)
```

```{r}
# Find predictor with minimum p-value (excluding intercept)
p_vals   <- coef_table[-1, "Pr(>|z|)"]           # drop intercept
min_pred <- names(which.min(p_vals))
min_pval <- min(p_vals)
min_coef <- coef_table[min_pred, "Estimate"]

cat("Significant predictors (alpha = 0.05):",
    ifelse(any(p_vals < 0.05),
           paste(names(p_vals[p_vals < 0.05]), collapse = ", "),
           "None"), "\n\n")

cat(sprintf("Predictor with the smallest p-value : %s\n", min_pred))
cat(sprintf("p-value  : %.4f\n", min_pval))
cat(sprintf("Estimated coefficient : %.4f\n", min_coef))
cat(sprintf("Odds ratio exp(coef) : %.4f\n", exp(min_coef)))

```
# Part (c): Odds of Market Going "Up" at Median Predictor Values

```{r}
# Compute median values of all predictors in the training data
med_vals <- train %>% summarise(across(c(Lag1, Lag2, Lag3, Lag4, Lag5, Volume), median))

cat("Median values of predictors:\n")
print(med_vals)

# Predict log-odds at median values
log_odds_med <- predict(logit_model, newdata = med_vals, type = "link")
prob_med     <- predict(logit_model, newdata = med_vals, type = "response")
odds_med     <- exp(log_odds_med)

cat(sprintf("\nLog-odds at median predictors : %.4f\n", log_odds_med))
cat(sprintf("Odds at median predictors : %.4f\n", odds_med))
cat(sprintf("Probability at median predictors : %.4f\n", prob_med))
```


Part (e): Optimal Cut-Point on Test Data (Minimising TPR × (1 − FPR))

We use the fitted model to predict probabilities on the test data, then find the threshold that **minimises** $TPR \times (1 - FPR)$ (equivalently, maximises $TPR \times FPR$, i.e., the Youden Index point inverted — we minimise the product of sensitivity and specificity complement).

```{r}
# Predicted probabilities on test data
test_probs <- predict(logit_model, newdata = test, type = "response")
test_dir   <- ifelse(test$Direction == "Up", 1, 0)   # 1 = Up, 0 = Down

# Compute TPR and FPR across a grid of cut-points
thresholds <- seq(0.01, 0.99, by = 0.001)

results <- sapply(thresholds, function(t) {
  pred_class <- ifelse(test_probs >= t, 1, 0)
  TP <- sum(pred_class == 1 & test_dir == 1)
  FP <- sum(pred_class == 1 & test_dir == 0)
  TN <- sum(pred_class == 0 & test_dir == 0)
  FN <- sum(pred_class == 0 & test_dir == 1)

  TPR <- ifelse((TP + FN) > 0, TP / (TP + FN), 0)   # Sensitivity
  FPR <- ifelse((FP + TN) > 0, FP / (FP + TN), 0)   # 1 - Specificity

  c(TPR = TPR, FPR = FPR, obj = TPR * (1 - FPR))
})

results_df   <- as.data.frame(t(results))
results_df$threshold <- thresholds

# Optimal threshold: minimise TPR * (1 - FPR)
opt_idx   <- which.min(results_df$obj)
opt_cut   <- results_df$threshold[opt_idx]
opt_TPR   <- results_df$TPR[opt_idx]
opt_FPR   <- results_df$FPR[opt_idx]

cat(sprintf("Optimal cut-point            : %.3f\n", opt_cut))
cat(sprintf("TPR at optimal cut-point     : %.4f\n", opt_TPR))
cat(sprintf("FPR at optimal cut-point     : %.4f\n", opt_FPR))
cat(sprintf("TPR*(1-FPR) at optimal cut   : %.6f\n", opt_TPR * (1 - opt_FPR)))
```

## Confusion Matrix at Optimal Cut-Point

```{r}
# Classify test observations using optimal cut-point
pred_opt <- factor(ifelse(test_probs >= opt_cut, "Up", "Down"),
                   levels = c("Down", "Up"))
actual   <- factor(test$Direction, levels = c("Down", "Up"))

# Confusion matrix (rows = Predicted, cols = Actual) — force both levels
conf_mat <- table(Predicted = pred_opt, Actual = actual)
print(conf_mat)

# Safe cell extraction (returns 0 if row/col missing)
safe_cell <- function(mat, r, c) {
  if (r %in% rownames(mat) && c %in% colnames(mat)) mat[r, c] else 0
}

TP <- safe_cell(conf_mat, "Up",   "Up")
TN <- safe_cell(conf_mat, "Down", "Down")
FP <- safe_cell(conf_mat, "Up",   "Down")
FN <- safe_cell(conf_mat, "Down", "Up")

total      <- sum(conf_mat)
test_error <- (FP + FN) / total
accuracy   <- 1 - test_error
sensitivity <- TP / (TP + FN)
specificity <- TN / (TN + FP)

cat(sprintf("\nTotal test observations : %d\n", total))
cat(sprintf("Correctly classified    : %d\n", TP + TN))
cat(sprintf("Misclassified           : %d\n", FP + FN))
cat(sprintf("Test Error Rate         : %.4f (%.2f%%)\n", test_error, test_error * 100))
cat(sprintf("Accuracy                : %.4f (%.2f%%)\n", accuracy,   accuracy   * 100))
cat(sprintf("Sensitivity (TPR)       : %.4f\n", sensitivity))
cat(sprintf("Specificity             : %.4f\n", specificity))
```

# Part (f): Model with Only Lag1 and Lag2

We refit the logistic regression using only `Lag1` and `Lag2` and compare the test error rate.

```{r}
# Fit reduced model (Lag1 + Lag2 only)
logit_reduced <- glm(Direction ~ Lag1 + Lag2,
                     data   = train,
                     family = binomial(link = "logit"))

summary(logit_reduced)
```

```{r}
coef_r <- as.data.frame(summary(logit_reduced)$coefficients)
coef_r$Variable <- rownames(coef_r)
rownames(coef_r) <- NULL
coef_r <- coef_r[, c("Variable","Estimate","Std. Error","z value","Pr(>|z|)")]
coef_r %>%
  mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
  kable(caption = "Reduced Model Coefficients (Lag1 + Lag2 only)",
        align = c("l","r","r","r","r")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE)
```

## Optimal Cut-Point for Reduced Model

```{r}
# Predictions on test data from reduced model
test_probs_r <- predict(logit_reduced, newdata = test, type = "response")

results_r <- sapply(thresholds, function(t) {
  pred_class <- ifelse(test_probs_r >= t, 1, 0)
  TP <- sum(pred_class == 1 & test_dir == 1)
  FP <- sum(pred_class == 1 & test_dir == 0)
  TN <- sum(pred_class == 0 & test_dir == 0)
  FN <- sum(pred_class == 0 & test_dir == 1)
  TPR <- ifelse((TP + FN) > 0, TP / (TP + FN), 0)
  FPR <- ifelse((FP + TN) > 0, FP / (FP + TN), 0)
  c(TPR = TPR, FPR = FPR, obj = TPR * (1 - FPR))
})

results_r_df <- as.data.frame(t(results_r))
results_r_df$threshold <- thresholds

opt_idx_r <- which.min(results_r_df$obj)
opt_cut_r <- results_r_df$threshold[opt_idx_r]

cat(sprintf("Optimal cut-point (reduced model): %.3f\n", opt_cut_r))

# Confusion matrix for reduced model — force both levels
pred_opt_r <- factor(ifelse(test_probs_r >= opt_cut_r, "Up", "Down"),
                     levels = c("Down", "Up"))
conf_mat_r <- table(Predicted = pred_opt_r, Actual = actual)
print(conf_mat_r)

TP_r <- safe_cell(conf_mat_r, "Up",   "Up")
TN_r <- safe_cell(conf_mat_r, "Down", "Down")
FP_r <- safe_cell(conf_mat_r, "Up",   "Down")
FN_r <- safe_cell(conf_mat_r, "Down", "Up")

test_error_r <- (FP_r + FN_r) / total
cat(sprintf("\nTest Error Rate (Lag1+Lag2 model): %.4f (%.2f%%)\n",
            test_error_r, test_error_r * 100))
```

```{r}
comparison <- data.frame(
  Model         = c("Full Model (Lag1–Lag5 + Volume)", "Reduced Model (Lag1 + Lag2)"),
  `Cut-Point`   = c(round(opt_cut,   3), round(opt_cut_r, 3)),
  `Test Error`  = c(paste0(round(test_error   * 100, 2), "%"),
                    paste0(round(test_error_r * 100, 2), "%")),
  check.names = FALSE
)
comparison %>%
  kable(caption = "Comparison of Test Error Rates", align = c("l","c","c")) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```





























