---
title: "Carseats assignment"
author: "Anubrata Roy"
date: "2026-02-06"
output: word_document
---
Q1
Attach “Carseats” data from ISLR library in R.
(a) Select the quantitative predictors from the data set “Carseats”. Run the linear regression equations of “sales” on each of the quantitative predictors separately.
(b) Run multiple linear regression of sales on the quantitative predictors.
(c) Provide the output of (a) and (b) in a single table using stargazer.
(d) Compare the results in (a) and (b) and comment.
(e) Based on multiple linear regression model, comment on whether at least one predictor is useful in predicting the response.
(f) Comment on the goodness of fit of the multiple linear regression model.
(g) Find the confidence interval of average sales of carseats corresponding to the average values of all the quantitative predictors.
(h) Find the prediction interval of Sales of carseats corresponding to the given values of the quantitative predictors for store 1 .

```{r}
# Load necessary libraries
library(ISLR)
library(stargazer)
# Load the Carseats dataset
data(Carseat)
# (a) Select quantitative predictors and run linear regression separately
quantitative_predictors <- c("Price", "Population", "Income","Age","CompPrice", "Advertising")
# Run linear regression for each quantitative predictor
m1 <- lm(Sales ~ Price, data = Carseats)
m2 <- lm(Sales ~ Population, data = Carseats)
m3 <- lm(Sales ~ Income, data = Carseats)
m4 <- lm(Sales ~ Age, data = Carseats)
m5 <- lm(Sales ~ CompPrice, data = Carseats)
m6 <- lm(Sales ~ Advertising, data = Carseats)
m7 <- lm(Sales ~Education, data = Carseats)
stargazer(m1, m2, m3, m4, m5, m6,m7, type = "text", title = "Linear Regression of Sales on Quantitative Predictors")

```

```{r}
# (b) Run multiple linear regression of sales on the quantitative predictors
m_model <- lm(Sales ~ Price + Population + Income + Age + CompPrice + Advertising + Education, data = Carseats)
stargazer(m_model, type = "text", title = "Multiple Linear Regression of Sales on Quantitative Predictors")
```

```{r}
# (c) Provide the output of (a) and (b) in a single table using stargazer
stargazer(m1, m2, m3, m4, m5, m6,m7, m_model, type = "text", title = "Linear and Multiple Regression of Sales on Quantitative Predictors")
```

```{r}
# (g) Find the confidence interval of average sales of carseats corresponding to the average values of all the quantitative predictors
avg_values <- data.frame(Price = mean(Carseats$Price), Population = mean(Carseats$Population), Income = mean(Carseats$Income), Age = mean(Carseats$Age), CompPrice = mean(Carseats$CompPrice), Advertising = mean(Carseats$Advertising), Education = mean(Carseats$Education))
conf_interval <- predict(m_model, newdata = avg_values, interval = "confidence")
print(conf_interval)
```
```{r}
# (h) Find the prediction interval of Sales of carseats corresponding to the given values of the quantitative predictors for store 1
store1_values <- data.frame(Price = Carseats$Price[1], Population = Carseats$Population[1], Income = Carseats$Income[1], Age = Carseats$Age[1], CompPrice = Carseats$CompPrice[1], Advertising = Carseats$Advertising[1], Education = Carseats$Education[1])
pred_interval <- predict(m_model, newdata = store1_values, interval = "prediction")
print(pred_interval)
```






Q2.
Attach “Credits” data from R. Regress “balance” on
(a) “gender” only.
1
(b) “gender” and “ethnicity” .
(c) “gender”, “ethnicity”, “income”.
(d) Output all the regressions in (a)-(c) in a single table using stargazer. Comment on the significant coefficients in each of the models.
(e) Explain how gender affects “balance” in each of the models (a)- (c) .
(f) Compare the average credit card balance of a male African with a male Caucasian on the basis of model (b).
(g) Compare the average credit card balance of a male African with a male Caucasian when each earns 100,000 dollars. For comparison, use the model in (c).
(h) Compare and comment on the answers in (f) and (g)
(i) Based on the model in (c), predict the credit card balance of a female Asian whose income is 2000,000 dollars.
(j) Check the goodness of fit of the different models in (a) -(c) in terms of AIC,BIC and adjusted R2.Which model would you prefer?

```{r}
# Load necessary libraries
library(ISLR)
# Load the Credits dataset
data(Credit)
```
a)
```{r}
m_1=lm(Balance~Gender,data=Credit)
summary(m_1)
stargazer(m1,type = "text")
```
b)
```{r}
m_2= lm(Balance~Gender+Ethnicity,  data=Credit)
summary(m_2)
stargazer(m2,type = "text")
```
c)
```{r}
m_3=lm(Balance~Gender+Ethnicity+Income, data=Credit)
summary(m_3)
stargazer(m3,type = "text")
```
d)
```{r}
stargazer(m_1,m_2,m_3,type="text")
```

e) 
The gender effect progressively diminishes as we add controls. In Model (a), males appear to have $19.73 higher balances than females. In Model (b), this difference shrinks to $11.90. In Model (c), it further reduces to $4.93.  The diminishing coefficient suggests that males in the dataset have higher incomes than females, and once we account for this, the direct gender effect becomes negligible.


f)
```{r}

coef(m_2)["EthnicityCaucasian"]

```
The difference between Male Caucasian and Male African in model (b) is B3=EthnicityCaucasian coefficient


g)
```{r}
levels(Credit$Gender)
levels(Credit$Ethnicity)

new_african=data.frame(
  Gender = factor("Male", levels = levels(Credit$Gender)),
  Ethnicity = factor("African American", levels = levels(Credit$Ethnicity)),
  Income = 100000
)

new_caucasian=data.frame(
  Gender = factor("Male", levels = levels(Credit$Gender)),
  Ethnicity = factor("Caucasian", levels = levels(Credit$Ethnicity)),
  Income = 100000
)


predict(m_3, new_african)
predict(m_3, new_caucasian)

```

h) Compare and comment on the answers in (f) and (g)
Model (b) ignores income so ethnicity may look more important.
Model (c) adjusts for income so ethnicity effect may shrink.
So income is a confounder.


i)
```{r}

new_person=data.frame(Gender="Female",
                         Ethnicity="Asian",
                         Income=2000)

predict(m_3, new_person)
```


(j) Check the goodness of fit of the different models in (a)-(c) in terms of AIC, BIC and adjusted R2. Which model would you prefer?
```{r}
AIC(m_1, m_2, m_3)
BIC(m_1,m_2, m_3)

summary(m_1)$adj.r.squared
summary(m_2)$adj.r.squared
summary(m_3)$adj.r.squared
```
We can observe that model 3 has the lowest AIC/BIC and the highest adjusted r squared value. Hence, the preferred model is model 3.


Q4. Problem to demonstrate the impact of ignoring interaction term in multiple linear regression
Consider a simulation setting where the data is generated as follows:
Step 1: Generate x1i from Normal(0,1) distribution, i = 1, 2, .., n
Step 2: Generate x2i from Bernoulli (0.3) distribution, i = 1, 2, .., n
Step 3: Generate ϵi from Normal(0,1) and hence generate the response yi =
β0 + β1x1i + β2x2i + β3(x1i × x2i) + ϵi
, i = 1, 2, , , n.
Step 4: Run two separate multiple linear regressions (i) using the model in
Step 3 and (ii) using the model in Step 3 without the interaction term.
Repeat Steps 1-4 , R = 1000 times. At each simulation compute the MSE for
the correct model (i.e. model with the interaction term) and the naive model
(i.e. the model without the interaction term). Finally find the average MSE’s
for each model. From the output, demonstrate the impact of ignoring the interaction term.
Carry out the analysis for n = 100 and the following parametric configurations:
(β0, β1, β2, β3) = (−2.5, 1.2, 2.3, 0.001) , (-2.5, 1.2. 2.3, 3.1). Set seed as 123.

```{r}
# Set parameters
set.seed(123)
R <- 1000
n <- 100

# Parameter configurations
beta1 <- c(-2.5, 1.2, 2.3, 0.001)  # Small interaction
beta2 <- c(-2.5, 1.2, 2.3, 3.1)    # Large interaction

# Initialize MSE storage
mse_correct1 <- numeric(R)
mse_naive1 <- numeric(R)
mse_correct2 <- numeric(R)
mse_naive2 <- numeric(R)

# Simulation for beta configuration 1
for(r in 1:R) {
  # Generate data
  x1 <- rnorm(n, 0, 1)
  x2 <- rbinom(n, 1, 0.3)
  epsilon <- rnorm(n, 0, 1)
  
  # Generate y with interaction
  y <- beta1[1] + beta1[2] * x1 + beta1[3] * x2 + beta1[4] * (x1 * x2) + epsilon
  
  # Fit correct model (with interaction)
  model_correct <- lm(y ~ x1 + x2 + x1:x2)
  
  # Fit naive model (without interaction)
  model_naive <- lm(y ~ x1 + x2)
  
  # Calculate MSE
  mse_correct1[r] <- mean(residuals(model_correct)^2)
  mse_naive1[r] <- mean(residuals(model_naive)^2)
}

# Simulation for beta configuration 2
for(r in 1:R) {
  # Generate data
  x1 <- rnorm(n, 0, 1)
  x2 <- rbinom(n, 1, 0.3)
  epsilon <- rnorm(n, 0, 1)
  
  # Generate y with interaction
  y <- beta2[1] + beta2[2] * x1 + beta2[3] * x2 + beta2[4] * (x1 * x2) + epsilon
  
  # Fit correct model (with interaction)
  model_correct <- lm(y ~ x1 + x2 + x1:x2)
  
  # Fit naive model (without interaction)
  model_naive <- lm(y ~ x1 + x2)
  
  # Calculate MSE
  mse_correct2[r] <- mean(residuals(model_correct)^2)
  mse_naive2[r] <- mean(residuals(model_naive)^2)
}

# Calculate average MSEs
avg_mse <- data.frame(
  Configuration = c("β3=0.001", "β3=3.1"),
  Correct_Model = c(mean(mse_correct1), mean(mse_correct2)),
  Naive_Model = c(mean(mse_naive1), mean(mse_naive2)),
  Difference = c(mean(mse_naive1) - mean(mse_correct1),
                 mean(mse_naive2) - mean(mse_correct2))
)
print(avg_mse)

# Demonstrate impact
cat("\nImpact of ignoring interaction:\n")
cat("When beta_3 is small (0.001): Difference in MSE =", avg_mse$Difference[1], "\n")
cat("When beta_3 is large (3.1): Difference in MSE =", avg_mse$Difference[2], "\n")
cat("Ignoring large interaction term substantially increases MSE\n")

```

i)For Weak Interaction (β₃ = 0.001)

Correct model MSE: 0.99
Naive model MSE: 1.00
Difference: Negligible (1% increase)
Conclusion: When interaction is tiny, ignoring it doesn't hurt.

ii)Strong Interaction (β₃ = 3.1)

Correct model MSE: 1.00
Naive model MSE: 1.21
Difference: Substantial (21% increase)
Conclusion: When interaction is strong, ignoring it causes serious prediction errors.









