---
title: "p_set_4"
author: "anubrata roy"
date: "2026-02-19"
output: html_document
---

Q1.Problem to demonstrate multicollinearity
Consider the Credit data in the ISLR library. Choose balance as the response and Age, Limit and Rating as the predictors.

(a) Make a scatter plot of (i) Age versus Limit and (ii) Rating Versus Limit. Comment on the scatter plot.

(b) Run three separate regressions: (i) Balance on Age and Limit (ii) Balance on Age, Rating and Limit (iii) Balance on Rating and Limit. Present all the regression output in a single table using stargazer. What is the marked difference that you can observe from the output?

(c) Calculate the variance inflation factor (VIF) and comment on multicollinear-
ity.



```{r}
# Load required libraries
library(ISLR)
library(stargazer)
library(car)

# Load the Credit data
data(Credit)

# Check the structure of the data
str(Credit)
```
a)
```{r}
# Set up plotting area
par(mfrow = c(1, 2))

# (i) Age versus Limit
plot(Credit$Limit, Credit$Age, 
     xlab = "Limit", ylab = "Age",
     main = "Age vs Limit",
     pch = 19, col = "blue")
abline(lm(Age ~ Limit, data = Credit), col = "red", lwd = 2)

# (ii) Rating versus Limit
plot(Credit$Limit, Credit$Rating, 
     xlab = "Limit", ylab = "Rating",
     main = "Rating vs Limit",
     pch = 19, col = "blue")
abline(lm(Rating ~ Limit, data = Credit), col = "red", lwd = 2)

# Reset plotting parameters
par(mfrow = c(1, 1))

# Calculate correlations
cor_age_limit <- cor(Credit$Age, Credit$Limit)
cor_rating_limit <- cor(Credit$Rating, Credit$Limit)
cat("Correlation between Age and Limit:", cor_age_limit, "\n")
cat("Correlation between Rating and Limit:", cor_rating_limit, "\n")
```

Age vs Limit: Shows weak/no clear pattern, indicating low correlation

Rating vs Limit: Shows strong positive linear relationship, indicating high correlation

This suggests multicollinearity issues when including both Rating and Limit


b)
```{r}
# Run three regressions
# (i) Balance on Age and Limit
model1 <- lm(Balance ~ Age + Limit, data = Credit)

# (ii) Balance on Age, Rating and Limit
model2 <- lm(Balance ~ Age + Rating + Limit, data = Credit)

# (iii) Balance on Rating and Limit
model3 <- lm(Balance ~ Rating + Limit, data = Credit)

# Create a single table using stargazer
stargazer(model1, model2, model3,
          type = "text",
          title = "Regression Results",
          column.labels = c("Model 1", "Model 2", "Model 3"),
          dep.var.labels = "Balance",
          covariate.labels = c("Age", "Rating", "Limit"),
          omit.stat = c("f", "ser"),
          digits = 3)

# Calculate standard errors for comparison
summary_model1 <- summary(model1)
summary_model2 <- summary(model2)
summary_model3 <- summary(model3)

cat("\nModel 1 - Age coefficient SE:", summary_model1$coefficients[2, 2], "\n")
cat("Model 2 - Age coefficient SE:", summary_model2$coefficients[2, 2], "\n")
cat("Model 2 - Rating coefficient SE:", summary_model2$coefficients[3, 2], "\n")
cat("Model 2 - Limit coefficient SE:", summary_model2$coefficients[4, 2], "\n")
cat("Model 3 - Rating coefficient SE:", summary_model3$coefficients[2, 2], "\n")
cat("Model 3 - Limit coefficient SE:", summary_model3$coefficients[3, 2], "\n")
```

Standard errors increase dramatically in Model 2 (with all three predictors) for Rating and Limit

Coefficient signs may change or become unstable

Individual p-values may become insignificant even though the model overall is significant

Rating and Limit coefficients show high variance due to multicollinearity

c)
```{r}
# Calculate VIF for the full model (Model 2)
vif_values <- vif(model2)
cat("Variance Inflation Factors:\n")
print(vif_values)

# Interpretation
cat("\nInterpretation:\n")
cat("- VIF = 1: No correlation\n")
cat("- VIF between 1-5: Moderate correlation (acceptable)\n")
cat("- VIF > 5: High correlation (problematic)\n")
cat("- VIF > 10: Serious multicollinearity\n")

# Check correlation matrix of predictors
cor_matrix <- cor(Credit[, c("Age", "Rating", "Limit")])
cat("\nCorrelation Matrix:\n")
print(cor_matrix)
```

Q2.Problem to demonstrate the detection of outlier, leverage and influential points

Attach “Boston” data from MASS library in R. Select median value of owner-occupied homes, as the response and per capita crime rate, nitrogen oxides concentration, proportion of blacks and percentage of lower status of the population as predictors. The objective is to fit a multiple linear regression model of the response on the predictors. With reference to this problem, detect outliers, leverage points and influential points if any.

```{r}
# Load required libraries
library(MASS)
library(car)

# Load Boston data
data(Boston)

# Select variables
# Response: medv (median value of owner-occupied homes)
# Predictors: crim (per capita crime rate), nox (nitrogen oxides concentration),black (proportion of blacks),lstat (percentage of lower status population)

# Create subset
boston_subset <- Boston[, c("medv", "crim", "nox", "black", "lstat")]

# Fit multiple linear regression model
model_boston <- lm(medv ~ crim + nox + black + lstat, data = boston_subset)
summary(model_boston)

### 1. Outlier Detection
# Studentized residuals
student_resid <- rstudent(model_boston)

# Identify outliers (|studentized residual| > 3)
outliers <- which(abs(student_resid) > 3)
cat("Outliers (|studentized residual| > 3):\n")
print(outliers)
cat("\nStudentized residuals for outliers:\n")
print(student_resid[outliers])

# Plot studentized residuals
par(mfrow = c(2, 2))
plot(student_resid, 
     ylab = "Studentized Residuals",
     main = "Outlier Detection",
     pch = 19, col = "blue")
abline(h = c(-3, 3), col = "red", lty = 2)

# Identify points with |studentized residual| > 3
points(outliers, student_resid[outliers], 
       col = "red", pch = 19, cex = 1.5)
```

```{r}
# Calculate leverage values (hat values)
leverage <- hatvalues(model_boston)

# Calculate threshold (2p/n where p = number of predictors, n = sample size)
p <- length(coef(model_boston)) - 1  # number of predictors
n <- nrow(boston_subset)
leverage_threshold <- 2 * p / n

# Identify high leverage points
high_leverage <- which(leverage > leverage_threshold)
cat("High leverage points (h > 2p/n =", round(leverage_threshold, 4), "):\n")
print(high_leverage)
cat("\nLeverage values for high leverage points:\n")
print(leverage[high_leverage])

# Plot leverage values
plot(leverage, 
     ylab = "Leverage (Hat Values)",
     main = "Leverage Points Detection",
     pch = 19, col = "blue")
abline(h = leverage_threshold, col = "red", lty = 2)
points(high_leverage, leverage[high_leverage], 
       col = "red", pch = 19, cex = 1.5)
```

```{r}
# Cook's Distance
cooks_d <- cooks.distance(model_boston)

# Threshold (4/n)
cooks_threshold <- 4 / n

# Identify influential points
influential <- which(cooks_d > cooks_threshold)
cat("Influential points (Cook's D > 4/n =", round(cooks_threshold, 4), "):\n")
print(influential)
cat("\nCook's D values for influential points:\n")
print(cooks_d[influential])

# Plot Cook's Distance
plot(cooks_d, 
     ylab = "Cook's Distance",
     main = "Influential Points Detection",
     pch = 19, col = "blue")
abline(h = cooks_threshold, col = "red", lty = 2)
points(influential, cooks_d[influential], 
       col = "red", pch = 19, cex = 1.5)

# DFBETAS (influence on individual coefficients)
dfbetas_values <- dfbetas(model_boston)

# Find points with |DFBETAS| > 2/sqrt(n)
dfbetas_threshold <- 2 / sqrt(n)
influential_coef <- apply(abs(dfbetas_values) > dfbetas_threshold, 1, any)
cat("\nPoints influential on individual coefficients (|DFBETAS| > 2/sqrt(n) =", 
    round(dfbetas_threshold, 4), "):\n")
print(which(influential_coef))

# DFFITS (influence on fitted values)
dffits_values <- dffits(model_boston)
dffits_threshold <- 2 * sqrt(p / n)
influential_fitted <- which(abs(dffits_values) > dffits_threshold)
cat("\nPoints influential on fitted values (|DFFITS| > 2*sqrt(p/n) =", 
    round(dffits_threshold, 4), "):\n")

print(influential_fitted)
```














