---
title: "assgnment3_Q5"
author: "anubrata roy"
date: "2026-02-19"
output: html_document
---

Q5.Problem to demonstrate the utility of non-linear regression over linear regression

Get the fgl data set from “MASS” library.
(a) Considering the refractive index (RI) of “Vehicle Window glass” as the
variable of interest and assuming linearity of regression, run multiple linear regression of RI on different metallic oxides. From the p value, report which
metallic oxide best explains the refractive index.
(b) Run a simple linear regression of RI on the best predictor chosen in (a).
(c) Can you further improve the regression of the refractive index of “Vehicle
Window glass” on the predictor chosen by you in part (a)? Give the new fitted
model and compare its performance with the model in (b).


```{r}
# Load required libraries
library(MASS)
library(ggplot2)
library(gridExtra)

# Load the fgl dataset
data(fgl)
# View the structure of the dataset
str(fgl)

# Filter for Vehicle Window glass only
vehicle_glass <- subset(fgl, type == "Veh")
cat("Number of Vehicle Window glass samples:", nrow(vehicle_glass), "\n")
```

a)
```{r}
# Multiple linear regression of RI on metallic oxides
# The metallic oxides in the dataset are: Na, Mg, Al, Si, K, Ca, Ba, Fe
mlr_model <- lm(RI ~ Na + Mg + Al + Si + K + Ca + Ba + Fe, data = vehicle_glass)

# Summary of the model
summary_mlr <- summary(mlr_model)
print(summary_mlr)

# Extract p-values
p_values <- summary_mlr$coefficients[-1, 4]  # Remove intercept
p_values <- sort(p_values)
cat("\nP-values for metallic oxides (sorted):\n")
print(p_values)

# Find the best predictor (lowest p-value)
best_predictor <- names(which.min(p_values))
cat("\nBest predictor of RI based on p-value:", best_predictor)
cat("\nP-value for", best_predictor, ":", min(p_values), "\n")
```
Best predictor of RI based on p-value: Fe
P-value for Fe : 0.003616 


b)
```{r}
# Simple linear regression using the best predictor
slr_model <- lm(RI ~ vehicle_glass[[best_predictor]], 
                data = vehicle_glass)
names(slr_model$coefficients) <- c("(Intercept)", best_predictor)

# Summary of the simple linear model
summary_slr <- summary(slr_model)
print(summary_slr)

# Calculate RMSE for model comparison
rmse_slr <- sqrt(mean(summary_slr$residuals^2))
cat("\nRMSE for Simple Linear Regression:", rmse_slr, "\n")
cat("R-squared for Simple Linear Regression:", summary_slr$r.squared, "\n")

```
RMSE for Simple Linear Regression: 1.652736 
R-squared for Simple Linear Regression: 0.2097192 

c) 

Yes, we can improve the regression by trying nonlinear relationships since the relationship between refractive index and metallic oxides isn't always perfectly straight.

```{r}
# Create quadratic model (adds squared term)
quadratic_model <- lm(RI ~ Mg + I(Mg^2), data = vehicle_glass)

# Compare with simple linear model
linear_model <- lm(RI ~ Mg, data = vehicle_glass)

# Check if quadratic term is significant
summary(quadratic_model)
```


```{r}
# Plot both models to see the improvement
ggplot(vehicle_glass, aes(x = Mg, y = RI)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linetype = "dashed") +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(title = "Nonlinear (red) fits better than Linear (blue)",
       subtitle = "The curved line follows the data more closely")
```












